\documentclass[11pt,oneside%,draft%
]{memoir}

% --- Packages ----------------------------------------------

\usepackage[USenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage{IEEEtrantools}
\usepackage{verbatim}
\usepackage{tocloft}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{braket}
\usepackage[hyphens]{url}
\usepackage{makeidx}
\usepackage[colorlinks=true,urlcolor=blue,linkcolor=blue,linktocpage=true]{hyperref}

% --- Book appearance ---------------------------------------

%\setstocksize{57em}{37em}
%\settrimmedsize{57em}{37em}{*}
%\setlrmarginsandblock{5em}{*}{1}
%\setulmarginsandblock{5em}{*}{1}
%\setlength{\headsep}{1.33em}
%\setlength{\footskip}{2.5em}
%\setlength{\parindent}{0em}
%\setlength{\parskip}{0.6em}
%\fixpdflayout
%\checkandfixthelayout

\setstocksize{6in}{6in}
\settrimmedsize{6in}{6in}{*}

%\setstocksize{8.7in}{6in}
%\settrimmedsize{8.7in}{6in}{*}
\setlrmarginsandblock{0.8in}{*}{1}
\setulmarginsandblock{0.8in}{*}{1}
\setlength{\headsep}{0.215in}
\setlength{\footskip}{2.5em}
\setlength{\parindent}{0em}
\setlength{\parskip}{0.6em}
\fixpdflayout
\checkandfixthelayout

\makepagestyle{thphp}
\makeevenhead{thphp}{\thepage}{\rightmark}{\thepage}
\makeoddhead{thphp}{\thepage}{\rightmark}{\thepage}
\makeoddfoot{thphp}{}{{}}{}
\makeevenfoot{thphp}{}{{}}{}
\pagestyle{thphp}

\renewcommand{\cftdot}{}
\setlength\cftparskip{1pt}

% --- Mathematical notation ---------------------------------

% Environments
\newenvironment{eqna}{\begin{IEEEeqnarray*}{c}}{\end{IEEEeqnarray*}\ignorespacesafterend}
\newenvironment{eqnb}{\begin{IEEEeqnarray*}{rCl}}{\end{IEEEeqnarray*}\ignorespacesafterend}
\newenvironment{narration}{\begin{em}}{\end{em}}
\newcommand{\nimi}[1]{\IEEEyesnumber\label{#1}}
\renewenvironment{equation}{\sdfsdfsd}{\sdfsf}
\renewenvironment{align}{\sdfsfsd}{\sdfsd}

% Derivatives
\newcommand{\der}[2]{\frac{\dd#1}{\dd#2}}
\newcommand{\pder}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\cder}[2]{\frac{D #1}{D #2}}

% Symbols
\newcommand{\puoli}{\frac{1}{2}}
\newcommand{\yksi}{\mathfrak{1}}
\newcommand{\andd}{\qquad\textrm{and}\qquad}
\newcommand{\orr}{\qquad\textrm{or}\qquad}
\newcommand{\wheree}{\qquad\textrm{where}\qquad}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\ee}{\mathrm{e}}
\newcommand{\circc}{\tau}
\newcommand{\paika}{\mathfrak{s}}


\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\dvec}[1]{\dot{\vec{#1}}}
\newcommand{\ddvec}[1]{\ddot{\vec{#1}}}
\newcommand{\pvec}[1]{\primed{\vec{#1}}}

% Operators
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Det}{Det}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\reaaliosa}{Re}
\DeclareMathOperator{\imaginaariosa}{Im}
\renewcommand{\Re}{\reaaliosa}
\renewcommand{\Im}{\imaginaariosa}

\newcommand{\primed}[1]{\hat{#1}}
\newcommand{\ind}[1]{\mathfrak{#1}}
\newcommand{\arxivreference}[1]{\url{#1}}

% Differential geometry
\newcommand{\chris}[3]{\{{_{#1}}\!{^{#2}}\!{_{#3}}\}}
\newcommand{\tensy}[2]{#1^{#2}}
\newcommand{\tensa}[2]{#1_{#2}}
\newcommand{\tensay}[3]{#1_{#2}^{\phantom{#2}#3}}
\newcommand{\tensya}[3]{#1^{#2}_{\phantom{#2}#3}}

% Dialog characters
\newcommand{\hea}{\(\blacklozenge\)\;}
\newcommand{\heb}{\(\Game\)\;}

% Colors and indices
\definecolor{safi}{RGB}{0,100,100}
\definecolor{oranssi}{RGB}{255,128,0}
\newcommand{\coa}{{\color{black}\bullet}}
\newcommand{\cob}{{\color{oranssi}\bullet}}
\newcommand{\coc}{{\color{cyan}\bullet}}
\newcommand{\cod}{{\color{red}\bullet}}
\newcommand{\coe}{{\color{magenta}\bullet}}
\newcommand{\cof}{{\color{green}\bullet}}
\newcommand{\cog}{{\color{safi}\bullet}}
\newcommand{\coh}{{\color{yellow}\bullet}}
\newcommand{\coi}{{\color{blue}\bullet}}

% UNCOMMENT THE NEXT LINE FOR TRADITIONAL GREEK INDICES
%\renewcommand{\coa}{\alpha}\renewcommand{\cob}{\beta}\renewcommand{\coc}{\gamma}\renewcommand{\cod}{\delta}\renewcommand{\coe}{\mu}\renewcommand{\cof}{\nu}\renewcommand{\cog}{\rho}\renewcommand{\coh}{\sigma}\renewcommand{\coi}{\xi}

% UNCOMMENT THE NEXT LINE FOR THE TRADITIONAL CIRCLE CONSTANT 2 PI
%\renewcommand{\circc}{2\pi}

\begin{document}
\frontmatter
\begin{titlingpage}
	\begin{centering}
		\HUGE\textbf{Time}\\
		\vspace{0.4em}
		\normalsize\emph{by}\\
		\vspace{0.4em}
		\textsc{Konsta Kurki}\\
		\vspace{0.4em}
		\textsc{\today}\\
		%\vspace{5em}
		%\vfill
		\vspace{5em}
		\textbf{---work in progress---}\\
	\end{centering}
	\vfill

	Copyright {\textcopyright} 2015 Konsta Kustaa Kurki

	This work is licensed under a Creative Commons Attribution 4.0 International License. See \url{http://creativecommons.org/licenses/by/4.0/} for the license and \url{http://github.com/konstakurki/time} for the source material.
\end{titlingpage}
\chapter{Change log}
\begin{narration}
This is an incomplete list of additions and changes made since the beginning of the writing.
\end{narration}


\textsc{June 28, 2015}\, Initial commit to \url{github.com/konstakurki/time}

\newpage
\tableofcontents
\mainmatter
%\book{The classical picture}
%\part{The very elements}
\chapter{The beginning}
\hea Excuse me.

\heb Yes?

\hea Could you tell me about time?

\heb Hmm. You mean the stuff you read from the upper right corner of your iPhone?

\hea I do not have an iPhone. But if I had I think it would be that.

\heb Why would you like to hear about it?

\hea I know how to read a clock but that's where my understanding ends and wondering begins. Sometimes I think I flow in time. Next moment I think time flows around me. And then I think nothing flows---or everything.

\heb I certainly feel you! Time's a peculiar fellow. Did you notice what kind of words you just used? Like `sometimes' and `moment'.

\hea Yeah, I know, it's embarrassing.

\heb You cannot escape him! Or her.

\hea I know. I'ts just hard to communicate without referring to time. Does it mean that I don't understand my own words?

\heb Well, it may be so. But don't worry---you're no worse than any one of us.

\hea You mean there's no person on Earth who understands his own words?

\heb I think understanding is just a feeling. Peace, acceptance. But it may be just a flash in time. In \emph{time!}

\hea For me it is not so easy to accept things like the fact that now it is a different moment than a couple of minutes ago. And now it is again different.

\heb It surely is.

\heb But see, there is no brief answer to your question.

\hea I think I can take a long one.

\heb You also think I would give you such?

\hea Wouldn't you?

\heb Your eyes relly seem to be searching for something. Maybe you really want to think these things. Am I right or are you just hoaxing me?

\hea I would really, really want to hear.

\heb All right, here's the deal: as long as you want to keep going, I will guide you through the universe, towards time. Our path will be long and difficult. But if I see you're not really interested, then I'll leave you just there. Or maybe we can do something else. But time needs your perfect focus. Understand?

\hea Understand.

\heb And one more thing. We problably won't make it to the end of the path. That's because I haven't reached it. As far as I know no one has. All we can do is just wander around and build some theory. Beautiful, but imperfect theory. Can you accept that?

\hea Just give me something. I must get somewhere with this question. Please. I've asked for many people and all they say is something really silly like 'time does not exist'. If something makes me that anxious it definitely must be there!

\heb I like you. I can feel the pain inside you with all my neurons. Not that I liked that someone does not feel good, but you seem to care about things that people usually do not even notice.

\heb Do you have any background?

\hea Acually I do have something. I've studied foundations of mechanics. I know Newton's laws, the action principle and Hamilton's equations. And I also know some quantum mechanics.

\hea But it's all really on the level of doing calculations. In university they don't really care of understandin; it's just all about graduating and becoming a cool stuper string theorist or something.

\heb Good! Then we can fly over some less essential stuff. I promise we'll be careful with the important things.
\subsection{Sharpening the question}
\heb Tell me some important things of your life.

\hea My family. And friends. Freedom.

\heb Something more primitive.

\hea Hmmm \ldots birth.

\heb Excellent! The primitive elements we use in construing our lives are events. How would you order your birth, the day you learned to speak and the cutting of your umbilical cord?

\hea I guess I could order them in many ways. Of course there is the one way of ordering them, but probably you don't mean that.

\heb Very often the right direction is so obvious and ridiculously simple that you do not even dear to think of it. So go on.

\hea Ok. So first I was born and then my umbilical cord were cut. Then, after an exhaustive sequence of trials and errors I said my first word. Am I right?

\heb Yes you are! What I mean by `time' is the affair---whatever it is in its deepest---that creates this order. Or at least I don't know how to describe it more precisely. Do you agree?

\hea I guess I do.

\heb We can measure it easily, right? For example by whatching the corner of the stupid iPhone you mentioned.

\heb Yes we can. But there is some pretty complicated technology inside that device. There are also simpler methods.

\hea Of course. I think people have investigated how much time there is between two events for generations by counting for example how many summers or perhaps sunsets fits in between the events. I guess any repeating phenomenon would do it.

\heb That is indeed how each and every clock work. The `time that firs in between the events' is called time interval of the events and is usually denoted by \(\Delta t\), the big triangle, the Greek capital delta meaning interval and \(t\) time.

\heb If we count sunsets we get a bigger number than if we counted summers. But what is beautiful is that the relation between the frequencies of sunsets and summers never changes! It's always something like 365. If we take any two good clocks, by which I mean a device built for measuring time intervals, their relative paces remain always the same.

\hea That definitely makes sense. If your clock has same kind of second today than mine, I assume with no though that same is true also tomorrow. Does this mean that the time interval of two events is in some sense an absolute concept?

\heb Up to the time scale fixed by the clock, yes. We can of course convert time intervals read from different clocks to each other by knowing the relative paces of the clocks.

\heb It is useful to choose some reference time interval, for example day which is the interval of two successive midnights, or the second you mentioned, and express all intervals as multiples of it. Then each and every human being get the same number of days for the same two events.

\hea So the time interval of two arbitrary events measured in days---or in seconds---is an absolute, observer--independent number. Sounds very reasonable.

\heb Now we can form a mathematical model of time. Take the real number line \(\mathbb{R}\) and attach events on it in such a way that bigger numbers correspond to later times and so that the time interval of two events corresponds to the difference in the numbers attached to the events. A real number \(t_E\) attached to an event \(E\) is called a time coordinate.

\heb Time coordinate is of course unique. It depends on the unit of time we chose. We can also choose any event to label with \(t=0\), that is, we are free to choose the origin of time. The numbers zero and one are algrbraically special, but they do not have any physical significance; they are just artifacts of our choice to use \(\mathbb{R}\) as our mathematical model.

\hea Interesting.

\heb The real number line is topologically a continuum, which means that we can sensibly think of points as close as we wish. Physically that means that we can talk about time intervals as short as we want.

\hea Is that realistic? I mean that I can easily think of a year, a day, a second and even a blink of an eye, but where do we know that we can divide the blink as sharply as we wish?

\heb We don't. If nothing more, the real number line is a very convenient way to label events. And I guess we relly have a certain feeling of time being continuous; at least it does not feel granural but smooth.

\hea That's true.

\hea But wait a second. I watched \emph{Interstellar} a couple of weeks ago. In one scene the main character Cooper gives a wrist watch to his daughter Murph and explains that when he returns from his space voyage and they compare their watches, they will find out that the clocks are not anymore in sync, even though the clocks are identical. Was it crap or what?

\heb No it wasn't. Cooper is referring to a concept called proper time. It means the personal time experienced by an observer which in fact is different for Cooper and Murph. But Cooper flies in very exotic circumstances. Here, on Earth where we cannot move as fast as Cooper and the proper times of us all are pretty much the same. That effectively universal proper time is called Galilean or Newtonian time and is the time which we described with a real number \(t\).

\hea So now you tell me that the time interval was not absolute. Why did you try to lie to me?

\heb In our everyday regime it really is absolute. We do not notice any deviations, and for hundreds of years people really believed in its strick absoluteness. If you are patient, we will get to the relativity. Right now it would probably be too complicated to think of time intervals as subjective, but later we will see that actually the relativistic picture is astonishingly simple and beautiful.

\hea Okay. I accept that, so let's keep \(t\).
\section{Space}
\heb Could we measure intervals of some other physical, perhaps even more concrete, thing than time?

\hea You must mean distances in space.

\heb Yes I do. If you were about to measure say the distance between two stones, how would you do that?

\hea I would take a rope, tie some knots equidistantly on it, tighten it between the stones and count the knots in between.

\heb Excellent! I would have taken a meter stick but this is much more elegant. The process is very similar to measuring time; here the rope plays the role of a clock and the knot spacing sets up a distance scale.

\hea And like time intervals, also spatial distances are absolute up to a distance scale!

\heb Yes, at least in the regime of everyday phenomena.

\heb Space is very much like time, but there is a difference: in time you can only imagine to go backwards and forward; in space you can also go left, right, up and down.

\hea This sounds a bit stupid to me. I think that in time I cannot move back nor forward; time just passes. In space I can move, but it takes some time. I mean that `moving' means that location changes with time. It is ridiculous to say that time changes with time.

\heb Yeah, yeah. That's why I said that you can imagine. It is somewhat sloppy language. But the point is that space is three--dimensional while time has only one dimension.

\hea Yes. That's a difference.
\subsection{Some properties}
\heb What is the shortest path between two points in space?

\hea If I tighten a rope between the points, it of course settles down to a configuration in which there are as little rope as possible between the points. That's what tightening means. So the path the rope takes is the shortest path between the two points. The shortest path is a straight path.

\heb Excellent. This is one of the elementary properties of our space. The triangle inequality of elementary geometry can be seen as a special case of this fact.

\heb Let me take a stone and throw it in the air. Or you can do it.

\hea Ok.

\heb And watch carefully.

% kuva: kivi lentaa

\hea I can imagine that the rock draws a curve in space.

\heb Yes. If you wait for a time \(\Delta t\), the curve really is curved, but if we now decrease \(\Delta t\) down to an almost vanishing time interval \(\dd t\), we get only a small portion of the originall curve which looks practically straight. Let me denote that small arrow by \(\dd\vec{x}\).

\heb Such an arrow is called a vector. A vector is characterized by its length and the direction it points to.

\heb Now we can divide \(\dd\vec{x}\) by \(\dd t\) and get another vector
\begin{eqna}
    \der{\vec{x}}{t}\doteq\dvec{x},
\end{eqna}
the velocity. If we used \(\dd t/2\) instead of \(\dd t\), the particle would have made only half of \(\dd\vec{x}\). The velocity does not depend on the particular value of \(\dd t\) if it is small enough.

\heb You can think of the vectors \(\dd\vec{x}\) and \(\dvec{x}\) as being localized at somewhere on the trajectory of the baseball, but right now it is not important.

\hea Why the location does not matter?

\heb It is because we can parallel transport a vector, that is, move it while keeping its length and direction unchanged. If you don't trust your sense of direction, you can do it for example with the aid of a compass.

\hea Wait a minute. If I go around the magnetic north pole, the needle of the compass turns around. I wouldn't really trust even on a compass.

\heb True. But there is also another elegant and physically interesting way of doing the parallel transport.

\hea What is it?

\heb Take a wheel of a bicycle and put it to spin. You will notice that the spinning makes it difficult to change the direction of the axis. With fast enough spin and hyva ripustus the wheel can be used to keep track of a direction.

\hea Oh, you are talking about a gyroscope. They're interesting devices.

\heb We can add two vectors by moving, of course by parallel transport, the other arrow starting from the end of the other. The sum is the arrow starting from the beginning of the first vector and ending at the end of the second. It is obvious that vectors can also be multiplied by numbers.

\heb Vector spaces are well--known objects in mathematics, and if you are interested, you can find some details from for example the 2006? book by Riley, Hobson and Bence. Vectors are quite intuitive fellows, perhaps just because we see them in the physical world everywhere around us.
\section{Tensors}
\heb If there is a lamb in a room, the air is hot near the lamb. The temperature, let me denote it by \(\phi\), varies through space. Such a thing, a thing describable by a number at every point in space, is called a scalar field.

\heb We may imagine building surfaces of equal temperature in space. Can you imagine how does a surface of thirty degrees of celcius look like?

\hea The room temperature is less than thirty degrees, and the lamb is probably hotter. I guess the temperature increases somewhat smoothly, so if we close the lamb from any direction, at some point we must pass a point in which the temperature is exactly thirty degrees.

\hea Those points points form the surface of thirty degrees. It probably looks more or less like a sphere with the lamb in its center.

\heb Exactly. If we look close enought, the surfaces of equal temperature look flat, and are spaced equidistantly.

\heb Now think about two points close to each other, separated by \(\dd\vec{x}\). The temperature difference between them is \(\dd\phi\). If you draw a picture of the surfaces of equal temperatures and \(\dd\vec{x}\), it is clear that \(\dd\phi\) is equal to the number of the surfaces pierced by \(\dd\vec{x}\). We may write it as
\begin{eqna}
    \dd\phi=\nabla\phi\cdot\dd\vec{x}.
\end{eqna}

\hea I see.

\heb So the temperature field \(\phi(\vec{x})\) defines a function at each point in space. That function, called the gradient of the temperature and denoted by \(\nabla\phi\), takes \(\dd\vec{x}\) in and produces the difference \(\dd\phi\) between the points separated by \(\dd\vec{x}\).

% kuva: vektori percettaa gradientin

\heb If you watch the picture, you see that the gradient is linear in \(\dd\vec{x}\), that is,
\begin{eqna}
    \nabla\phi\cdot(a\,\dd\vec{x}+b\,\dd\vec{y})=a\nabla\phi\cdot\dd\vec{x}+b\nabla\phi\cdot\dd\vec{y}
\end{eqna}
with \(a\) and \(b\) numbers and \(\dd\vec{x}\) and \(\dd\vec{y}\) vectors.

\heb A function like \(\nabla\phi\), that is a linear function which takes a vector and produces a number, is called a one--form. If we have two one--forms \(\omega\) and \(\sigma\), we may define their linear combination as
\begin{eqna}
    (a\,\omega+b\,\sigma)\cdot\vec{x}=a\,\omega\cdot\vec{v}+b\,\sigma\cdot\vec{v}
\end{eqna}
for any numbers \(a\) and \(b\) and any vector \(\vec{v}\).

\hea So the one--forms also form a vector space

\heb Yes. It is called the dual vector space. Sometimes one--forms are called dual vectors.

\heb We can also consider linear maps, or tensors as they are usually called, which take any number of vectors as inputs and are linear in every input vector.

\heb The notation becomes a bit complex if a tensor has many inputs. That is why I think we should adopt a funny but practical notation. If we have for example a tensor \(T\) which takes three vectors and the input vectors are \(\vec{a}\), \(\vec{b}\) and \(\vec{c}\), we could write
\begin{eqna}
    T(\vec{a},\vec{b},\vec{c})\doteq\tensa{T}{\coa\cob\coc}\vec{a}^\coa\vec{b}^\cob\vec{c}^\coc.
\end{eqna}

\hea Looks hilarious. What are those colored balls?

\heb They are wireless connectors that plug the vectors into the tensor. People usually prefer greek letters like \(\alpha\) and \(\mu\), but I think the colored balls are more clear. If you would like to see letters instead of colors, you must go to the source code of our lives and uncomment one line.

\hea I like color.

\heb Me too! The order of writing of the vectors and the tensor does not matter, the colors keep track of which connector, or index as they are often called, is connected to which. A single vector \(\vec{v}\) mapped by a dual vector \(\omega\) can be therefore written as
\begin{eqna}
    \omega\cdot\vec{v}=\omega_\coa\vec{v}^\coa=\vec{v}^\coa\omega_\coa.
\end{eqna}
This suggests that we can think of \(\vec{v}\) as a tensor which takes the one--form \(\omega\) as its input! It does not really matter how we choose to think of it; we just connect an upper index to a lower and that's it.

\hea So we could have tensors with lower and upper indices, for example \(\tensya{K}{\coa\cob\coc}{\cod\coe\cof}\), taking vectors and dual vectors as inputs.

\heb Yes, we very well could. And as you probably guess, we can write any number of tensors in a row and then connect, or contract as it is often said, any lower index with any upper one.

\hea Actually it is useful to think that we first form an outer product, for example
\begin{eqna}
    \tensya{(K\otimes T)}{\coa\cob\coc}{\cod\coe\cof\cog\coh\coi}\doteq\tensya{K}{\coa\cob\coc}{\cod\coe\cof}\tensa{T}{\cog\coh\coi},
\end{eqna}
It is a tensor which takes vectors and dual vectors just as you would guess from the indices. It's value is the value of \(T\) fed with three vectors multiplied byt \(K\) fed by three vectors and three one--forms. It is easy to verify that it is linear in every index. Then we contract any upper index--lower index pairs we want.

\heb We can for example contract all three upper indices with the last three lower indices of the above outer product and get
\begin{eqna}
    \tensya{(K\otimes T)}{\coa\cob\coc}{\cod\coe\cof\coa\cob\coc}\doteq\tensa{(KT)}{\cod\coe\cof}
\end{eqna}
which is a tensor with three lower indices.

\hea How elegant is that! I just take some tensors, form an outer prodct of them by writing them in a row in whichever order I want, and then contract the indices I want. Every contraction removes one upper and one lower index. If there are equal numbers of upper and lower indices and I contract them all, I am left with a number, right?

\heb It is just that simple. Vectors are tensors with one upper index and one--forms tensors with one lower index. A number can be thought of as a tensor with no indices at all.

\hea Can I take a linear combination of any two tensors?

\heb If they have the same index structure, then yes. Tensors of a specific type form a vector space. So there is a vector space of tensors of type \(T_{\coa\cob}\), of type \(\tensya{T}{\coa}{\cob}\), of type \(\tensay{T}{\coa}{\cob\coc}\) and so on.

\hea This is very cool. I think I'm gonna read something about vector spaces and tensors.
%
%\heb Yeah, do that! One 
\subsection{Inner product}
\hea I read that a vector space is often equipped with an inner product. An inner product of two vectors is linear and symmetric in both of the vectors. I'm a bit confused, because Wikipedia denoted the outer product of two vectors \(\vec{v}\) and \(\vec{w}\) by \(\vec{v}\cdot\vec{w}\), that is exactly the same way as we denoted a vector mapped by a one--form.

\hea I guess that since the inner product is linear, we could write
\begin{eqna}
    \vec{v}\cdot\vec{w}=g_{\coa\cob}\vec{v}^\coa\vec{w}^\cob
\end{eqna}
with \(g\) a tensor.

\heb Great! Because the inner product is symmetric, we have \(g_{\coa\cob}=g_{\cob\coa}\). This may seem a bit sloppy, but it means that
\begin{eqna}
    g_{\coa\cob}\vec{v}^\coa\vec{w}^\cob=g_{\cob\coa}\vec{v}^\coa\vec{w}^\cob
\end{eqna}
for any vectors \(\vec{v}\) and \(\vec{w}\).

\heb An inner product describes the overlap of two vectors. If consider the overlap of a vector, for example a separation \(\dd\vec{x}\), with itself, we get a number wich is proportional to the square of the length of \(\dd\vec{x}\). So there is a natural way to choose the tensor \(g\).

\hea Let me guess: we choose \(g\) so that
\begin{eqna}
    |\dd\vec{x}|^2=g_{\coa\cob}\dd\vec{x}^\coa\dd\vec{x}^\cob
\end{eqna}
exactly?

\heb Yes. The tensor \(g\) is called the metric tensor of space, since it measures the distance separated by \(\dd\vec{x}\). For \(\dd\vec{x}\) one usually writes \(|\dd\vec{x}|^2\doteq\dd s^2\) and for any other vector \(\vec{v}\) that \(|\vec{v}|^2\doteq\vec{v}^2\). It is a bit inelegant to have different conventions for different vectors but this notation is standard and widespread.
\subsection{Rising and lowering}
\heb The metric is so important that for each vector \(\vec{v}^\coa\) it is useful to define a dual vector \(g_{\coa\cob}\vec{v}^\cob\doteq\vec{v}_\coa\). That is, we really do not bother to think vectors and dual vectors as different kinds of things; we only have vectors which can be written the index in the upstairs or the downstairs.

\hea Okay, now I see why the inner product was written in the same way as a vector mapped by a one--form.

\heb We can actually lower any upper index of any tensor in the same way, for example \(g_{\coa\cob}T^{\cob\coc}\doteq\tensay{T}{\coa}{\coc}\).

\hea You said that we don't bother to fundamentally distinguish vectors and dual vectors. So we can raise an index back up as easily as we lowered it?

\heb Yes we can. For this purpose we define an inverse metric \(g^{\coa\cob}\) by
\begin{eqna}
    g^{\coa\cob}g_{\cob\coc}=1^{\coa}_{\coc},
\end{eqna}
in which \(1\) is the unit tensor which does absolutely nothing. We could also think of \(1\) as the metric \(g\) with the other index rised. So first lowering and then raising is doing nothing as it should be. And the same holds of course for first rising and then lowering.

\hea So the picture simplifies even more. We write tensors with indices placed whereever it is practical and then contract. The metric never needs to be written explicitly. Damn this is elegant.

\heb Yeh, however, of course we may sometimes want to write it visible, for example when I write
\begin{eqna}
    \dd s^2=g_{\coa\cob}\dd\vec{x}^\coa\dd\vec{x}^\cob
\end{eqna}
I want to emphasize that the metric defines the distance of two nearby points.

\heb Note that rising and lowering were made possible by the existence of the metric. But our familiar space is just on example of a manifold, a highly useful mathematical concept. A manifold is kind of a smooth space with a definite dimension. Not ecery manifold has a distinguished metric. if so, then we may really need to distinguish between vectors and dual vectors.

\heb I should probably mention that sometimes we also do not have gyroscopes at our hands and it is not necessarily possible to parallel transport vectors far away. However, vectors located at the same point can still be added. This is because a manifold---actualy a differenitable manifold, to be precice---is smooth enough to allow parallel transporting vectors for small distances and therefore make moving short arrows one after other possible.

% bold used in space
\subsection{Length of a curve}
\heb Now we can calculate the length of any path or curve in space by
\begin{eqna}
    s=\int\dd s=\int\sqrt{g_{\coa\cob}\dd\vec{x}^\coa\dd\vec{x}^\cob}.
\end{eqna}
We can also parametrize the path for example by the elapsed time \(t\) of a body moves along the path. Then we have
\begin{eqna}
    s=\int\sqrt{g_{\coa\cob}\der{\vec{x}^\coa}{t}\der{\vec{x}^\cob}{t}}\,\dd t=\int\sqrt{g_{\coa\cob}\dvec{x}^\coa\dvec{x}^\cob}\,\dd t=\int|\dvec{x}|\,\dd t.
\end{eqna}
The first equality comes simply from the fact that
\begin{eqna}
    \frac{\dd t}{\sqrt{\dd t\,\dd t}}=1.
\end{eqna}
\subsection{Differentiating tensors}
\heb Vectors can be added together and multiplied by numbers, so we can differentiate a vector \(\vec{x}\) that depends for example on \(t\) as we differentiate any quantity,
\begin{eqna}
    \dot{\vec{v}}=\frac{\vec{v}(t+\dd t)-\vec{v}(t)}{\dd t},
\end{eqna}
and the result is a new vector. The variable \(t\) do not need to be time; it can be for example the position on some curve in space, if there is a vector in every point on the curve.

\heb We can do the same for any tensor. Like for almost any product, the Leibnitz rule holds also for outer products of tensors. For example
\begin{eqnb}
    \der{}{t}(\vec{v}\cdot\vec{w})&=&\der{}{t}\left(g_{\coa\cob}\vec{v}^\coa\vec{w}^\cob\right)\\
                       &=&\dot{g}_{\coa\cob}\vec{v}^\coa\vec{w}^\cob+g_{\coa\cob}\dvec{v}^\coa\vec{w}^\cob+g_{\coa\cob}\vec{v}^\coa\dvec{w}^\cob.
\end{eqnb}
If \(\vec{v}\) and \(\vec{w}\) are any unchanging vectors, also their inner product remains unchanged. Therefore \(\dot{g}\equiv0\). The metric tensor is constant and the Leibnitz rule for the inner product reads
\begin{eqna}
    \der{}{t}(\vec{v}\cdot\vec{w})=\dvec{v}\cdot\vec{w}+\vec{v}\cdot\dvec{w}.
\end{eqna}

\heb Because the metric is constant, it holds that
\begin{eqna}
    \dvec{v}_\coa=\der{}{t}(g_{\coa\cob}\vec{v}^\cob)=g_{\coa\cob}\dvec{v}^\cob,
\end{eqna}
that is, the derivative of a contraction is the contraction of a derivatie. This makes thing even more easy: we can just think of the derivative of a vector---or any other tensor---indices placed anywhere we want.

\hea Allright. Sounds reasonable. But why is Leibniz rule true in the first place? I mean if we have for example functions \(f(t)\) and \(g(t)\), then why
\begin{eqna}
    \der{fg}{t}=\dot{f}\,g+f\,\dot{g}?
\end{eqna}

\heb That's a good question. For the change \(\dd(fg)\) it holds that
\begin{eqna}
    \dd(fg)=(f+\dd f)(g+\dd g)-fg=\dd f\,g+f\,\dd g+\dd f\,\dd g.
\end{eqna}
Because the changes are small, the last thing is negligible. After dividing by \(\dd t\) we get just what you wrote. As you can see, this reasoning is quite general and does not really care were \(f(t)\) and \(g(t)\) numbers, tensors or whatever.

\heb If we have tensor field, for example a vector \(\vec{v}(\vec{x})\) at every point in space, then the difference \(D\vec{v}\) of vectors at points separated by \(\dd\vec{x}\) is clearly linear in? \(\dd\vec{x}\). We can therefore define a tensor \(\nabla\vec{v}\), called covariant derivative, in such a way that
\begin{eqna}
    D\vec{v}^\coa=\nabla_\cob\vec{v}^\coa\dd\vec{x}^\cob
\end{eqna}
for any \(\dd\vec{x}\).

% note: x is not a vector

\hea Why did you write \(D\vec{v}\) instead of \(\dd\vec{v}\)?

\heb It is due to subleties that become important when we use numbers to represent tensors. We may talk about it later.

\hea Ok. I presume that we can do the same for any tensor field and write for example
\begin{eqna}
    DT_{\coa\cob\coc}=\nabla_\cod T_{\coa\cob\coc}\dd x^\cod.
\end{eqna}

\heb Yes we can. If the point we consider moves with time, we have we can divide by \(\dd t\) and get
\begin{eqna}
    \dot{T}_{\coa\cob\coc}=\frac{D\vec{v}^\coa}{\dd t}=\frac{\nabla_\cod T_{\coa\cob\coc}\dd\vec{x}^\cod}{\dd t}=\nabla_\cod T_{\coa\cob\coc}\dvec{x}^\cod.
\end{eqna}

\heb It looks stupid to have different kinds of d's in the same expression.

\hea Yeah, maybe we should just write
\begin{eqna}
    \dot{T}=\frac{D T}{Dt}.
\end{eqna}

\heb Note that differentiating a tensor field requires comparing tensors located at nearby points, which can only be achieved by parallel transport. If the manifold does not have a parallel transport, then the covariant derive cannot be formed.

\hea But we just added vectors by parallel transpoprting them short distances. You said that it can be done because the manifold is smooth.

\heb Yeah, but if we differentiate tensors, the nearby tensors are nerly the same, and their difference is already small. It is of the same order of smallness that the error that comes from parallel transposrting by just using the smoothness and cannot be neglected.??
\section{Perspectives}
\heb Things don't look the same from the perspectives of different persons, right?

\hea Definitely not. A capitalist feels envy for a poor guy getting an unemployment compensation. The capitalist thinks that the poor guy hasn't done anything to earn it. On the other hand the poor guy thinks that the capitalist is acting selfishly since he has so much money that he could save many from starvation but is not willing to do that.

\heb Yeah, that's so very absurd. Even though we have all the technology to give everyone a good life and make hating old--fashioned, we're not willing to pursue that opportunity.

\heb But there are also another kinds of differences of perspectives that are more relevant for us here.

\hea I know, I just wanted to make an important remark. The motion of a rollercoaster looks different when observed from different locations.

\heb Also the look changes if the observer turns his eyes and looks to different locations. But we are smart enough to view space as a somewhat objective thing and can think of things like locations and vectors without choosing any particular observer positioned and orientated in a particular way, right?

\hea Yeah.

\heb But there is an affair that matters.

\hea What is it?

\heb Motion. The differences in the motions of observers cannot be dealt so easily as the differences in positions and orientations.

\hea Why is that? We just have to take into account that locations change due time. If I have two points and their separation \(\dd\vec{x}\), a moving observer sees both points moving and understans that the separation \(\dd\vec{x}\) does not change at all. Both observers can understand it as the same separation vector. I would like to that think vectors and tensor are the same even though people may move with respect to each other.

\heb Well, for the separation vector of two fixed points, it really is so. Also the metric, which represents the overlap of two vectors can be seen as the same, because the overlap of two vectors must remain unchanged if the vectors itself remain unchanged.

\heb But think about the velocity of a particle. One observer sees its position changing by \(\dd\vec{x}\) in a time interval \(\dd t\). But another observer, moving with respect to the first, sees a change \(\dd\pvec{x}=\dd\vec{x}+\dd\vec{o}\) in which \(\dd\vec{o}\) is the motion of the first observer from the perspective of the first. For the velocity vectors \(\dvec{x}\) and \(\dot{\pvec{x}}\) it holds that
\begin{eqna}
    \dot{\pvec{x}}=\der{\pvec{x}}{t}=\der{\vec{x}+\dd\vec{o}}{t}=\dvec{x}+\dvec{o}.
\end{eqna}
We cannot think of the velocity vector objectively, unless we assume that there is some natural state of `resting'.

\hea Do we?

\heb Aristotle? and co. assumed that natural resting is not moving with respect to Earth.

\heb But think about some simple phenomena, for example a swinging pendulum. Does the swinging care about motion?

\hea Well, if I put the pendulum on top of a car the air probably disturbs it, but yeah, I get the point. Inside the car it swings just like inside a house standing firmly on rock, at least if the ride is not so bumpy.

\heb Correct. If we eliminate disturbances like wind, the pendulum siwings the same way always when the lab is moving with a constant velocity. A house, a uniformly moving car, an elevator are all equally good labs.

\heb The idea that there is a natural equality of observers that move uniformly with respect to each other is called the principle of relativity. It seems to be very generally true.

\heb However, accelerating observers, I mean those that are moving nonuniformly with each other, are not equal. Pendulum swings differently in carousel and in a car driven by a mad man.

\hea Interesting. And sensible.

\heb If two observers move uniformly, with a constant velocity \(\dvec{o}\) with respect to each other, we have \(\dot{\pvec{x}}=\dvec{x}+\dvec{o}\) as we noted. But for the acceleration \(\ddot{\vec{x}}\) it holds that
\begin{eqna}
    \ddot{\pvec{x}}=\der{}{t}\dot{\pvec{x}}=\der{}{t}(\dvec{x}+\dvec{o})=\ddot{\vec{x}}
\end{eqna}
since \(\ddot{\vec{o}}\equiv 0\). The principle of relativity could be loosely formulated by saying that \emph{acceleration is absolute, velocity is not.}
\subsection{Newton's first law}
\heb Now we are ready to formulate an equation describing a law of nature.

\hea Cool!

\heb Consider a body, for example a/the stone, isolated as well as we can from the rest of the world. For simplicity let us neglect the possible rotation of the rock, that is, we think of it as being a point particle.

\hea Well, I think we need to eliminate the resistance of air. But rocks are quite heavy objects compared to their size, so the air resistance really does not matter. Another thing we should eliminate is gravity. But I don't know how we could do that. do you?

\heb Yes and no! You asked a question that will take us very far. Really. But let us not go there yet. If we cannot eliminate gravity, the stone makes an arc. But if the stone moves really fast and we consider only a very small time scale, then gravity does not really have time to have any effect and we may neglect it. Or, we may imagine soomehow compensating the gravitational force with some other force. Or maybe we could just remove gravity somehow.

\hea Okay, let us assume there is no gravity. The the stone would probably go just straight.

\heb Yes it would. According to the principle of relativity, the equation describing the motion of the stone, or any free particle, cannot involve the velocity directly. If we have neglected the effect of the environment, even gravity, then the law describing the stone cannot involve the location of the particle directly and must be independent of the direction where the stone is going. These notions are usually called the homogenity and isotropy of space.

\heb So, write on the blackboard the simplest law, consistent with these requirements, that you can ever imagine!

\heb Come on, be brave.

\hea Hmm.
\begin{eqna}
    \ddot{\vec{x}}=0
\end{eqna}

\heb \emph{Voil\`a!} You just invented Newton's first law: free particles move straight, with unchanging velocity.

\hea There are also another equations consistent with our requirements, for example \(\dddot{\vec{x}}=0\). But that would mean that the particle did somehow remember its acceleration. It would also mean that even acceleration were not absolute, only the change of acceleration, or jerk. But our experience tells us that acceleration is absolute. Newton's first law is in great agreement with experiments.

\hea How could it be that the simplest answer tend to be the correct one? I mean, I do really feel that `If I know nothing obout it, what else would I guess', but it still feels a bit, or not even just a bit, muddy.

\heb Say no more. In philosophy its called \emph{Occam's razor} and in some areas of physics the \emph{minimal substitution principle}. I think it is just an integral part of common sense. Impossible to prove or argue for, but also impossible to live without.

\heb It is somewhat similar issue with the issue of probability theory. Why do you expect that out of very many throws of dice about one sixth are sixes?

\hea I have no idea. Or I would say that because of summetry. But I cannot say why symmetry of the dice implies symmetry of the probabilities.

\heb Exactly. It just feel obvious. Even the rigorous mathematical theory of probabilities and measures and stochastics just lean on that asumption.

\hea But it is a very good assumption. What is the chances that it is wrong?

\hea Wait. What did I just say? Probabilities seem to be as enigmatic as time.

\heb Yeah they does. But if we want to focus on time, let's move on. Maybe we can come back to probabilities at some point?.

\hea Yeah, let's do that.
\chapter{Some physics}
\heb If you had to affect the motion of the stone somehow, what would you do?

\hea I could for example push it.

\heb Yeah. You could generate force with your muscles and direct it with your limbs. Let us denote the force you excert on the stone by \(\vec{F}\). It is pretty obvious that force is a vector: it has a direction and a magnitude. If the stone is otherwise free, how does it react to the force?

\hea If Newton's first law says that the velocity of the free particle remains constant, maybe the force changes it.

\heb Yes, yes, yes! Here it comes again, Occam's razor or just plain common sense, and writes
\begin{eqna}
    \vec{F}=\ddot{\vec{x}}.
\end{eqna}
Is this good? At least it reduces to the first law if \(\vec{F}=0\).
\section{Newton's second law}
\hea Different stones, for example a small and a big one, react differently to a force.
% motion is continuous and smooth

\heb Right, so let us add a constant \(m\), called mass, which quantifies that and write
\begin{eqna}
    \vec{F}=m\ddot{\vec{x}}.
\end{eqna}
This equation is called Newton's second law. The greater the mass of the particle, the more resistant it is for the changes in its motion.

\hea That seems reasonable. But what about \(\vec{F}\), is it clear that it is absolute in the sense that the acceleration is but the velocity is not?

\heb No. But it is also not perfectly clear how to attach an arrow describing the force that comes out of your muscles. At least it is not as easy as drawing an arrow between two fixed points in space.

\hea Here we need to decide how to define things. We need to guess what would be a reasonable definition, and here it is not perfectly clear. Let us assume that the force \emph{is} absolute in the sense in which the acceleration is, and let us use the second law as a definition of force.

\hea Could different definitions lead to different physical results?

\heb Yes they could. That is because the force is something concrete, something that you relly produce with your body. Newton's secobnd law really says that the mass is the only way in which bodies reactions to force differ.

\heb We could for example let the mass depend on the acceleration. And that dependence could be different for different particles. We just guess the Newton's second.

\hea I'm not perfectly comfortable with this reasoning.

\heb I understand you, but the situation could be so much worse! Think about some theories of psychology.

\hea Are you dissing psychology?

\heb No, I'm just pointing out how lucky we are. Our task is so much easier.

\heb And I can tell you that Newton's second law is pretty good a guess---it is valid at very large range of phenomena.
\subsection{Particle in a gravitational field}
\heb We already considered a simple example, the stone thrown up in the air. If we throw the stone straight up, its motion is vertical and we need only one number, say \(h\), to describe the location. We can take \(h\) to be the distance from the stone to the ground.

\heb Could you write down the equation of motion?

\hea Well, I guess the only relevant force in the problem is the gravitational force. It points straight down. Let me call it \(G\). The equation of motion thus reads
\begin{eqna}
    G=m\ddot{h}.
\end{eqna}

\heb Yes. Let us scale the unit of force so that \(G\rightarrow mG\) and the equation of motion becomes just \(G=\ddot{h}\).

\heb Does \(G\) depend on anything?

\hea Gravity feels the same on the sea level and on the top of a mountain. So \(G\) probably doesn't depend on \(h\) in any significant way. But I don't know why couldn't it depend on for example \(\dot{h}\).

\heb If you think about it, gravity feels the same in a steady and in a moving elevator, at least if the elevator is not accelerating. So the gravitational force seems to respect the principle of relativity.

\hea Okay, I see. If the \(G\) depended on \(\dot{h}\), the equation of motion would involve the velocity directly and would not be consistent with the principle of relativity. So \(G\) is constant.

% pitaa selittaa mita massa on

\heb Now we have here a very simple differential equation: the second derivative of \(h\) is constant. A second degree polynomial has exactly this property. So what do you do?

\hea I take
\begin{eqna}
    h(t)=at^2+bt+c
\end{eqna}
and substitute it to the equation. I get \(G=2a\). So \(a\) must be equal to \(G/2\) and \(b\) and \(c\) can be anything. This fits to my intuition according to which we need to specify the initial velocity and location to know the future motion.

\heb Perfect!
\subsection{Kinetic energy}
\heb In physics there is a concept called work, denoted by \(W\). What does work bring into your mind?

\hea Thinking. Pain, tiredness, but also satisfaction. But yeah, that is probably not relevant here. I also recognize the old--fashionend `blood, sweat and tears' stylee work, for example lifting sldfjsdlf .

\heb That old--fashioned work is physical work. If you for example raahaa a heavy stone on asphalt, the work done is proportional to the distance the stone moves and to the force needed for dragging. So, at least if the particle moves in the same direction than the force points, we have \(\dd W\propto\vec{F}\cdot\dd\vec{x}\).

\hea If the particle moves perpendicularly to the force, then I don't really do work; I just maintain the force. For example if I stand besides a railroad and push with a scateboard a train that passes along, the wheels of the scateboard making sure I don't get disturbed bu the motion of the train, I don't really do anything. I just push. So what you wrote seems reasonable even if the force and motion are not perpendicular to each other.

\heb Yes. Let us define \(\dd W=\vec{F}\cdot\dd\vec{x}\).

\heb If you push some object with a constant force and it moves with a constant velocity, we have
\begin{eqna}
    \dot{W}=\der{}{t}(\vec{F}\cdot\dd\vec{x})=\vec{F}\cdot\der{\vec{x}}{t}=\vec{F}\cdot\dvec{x},
\end{eqna}
so the greater the velocity, the greater \(\dot{W}\), or power, is needed. This is in accordance with everyday observations: the faster you drag the stone, the more you sweat.

\heb Now think of accelerating a particle of mass \(m\) from rest to the velocity \(\dvec{x}=\vec{v}\), for example giving a stone some velocity when throwing it. The work that we need to do, denoted in this context by \(T\), is
\begin{eqna}
    T=\int_0^\vec{v}\dd W=\int_0^\vec{v}\vec{F}\cdot\dd\vec{x}.
\end{eqna}
Newton's second law tells that \(\vec{F}=m\ddot{\vec{x}}\), so
\begin{eqna}
    T=\int_0^\vec{v} m\der{\dvec{x}}{t}\cdot\dd\vec{x}=m\int_0^\vec{v}\dd\dvec{x}\cdot\der{\vec{x}}{t}=m\int_0^\vec{v}\dvec{x}\cdot\dd\dvec{x}.
\end{eqna}
Leibniz rule tells us that
\begin{eqna}
    \dd(\dvec{x}^2)=\dd(\dvec{x}\cdot\dvec{x})=\dd\dvec{x}\cdot\dvec{x}+\dvec{x}\cdot\dd\vec{x}=2\dvec{x}\cdot\dd\dvec{x}
\end{eqna}
so we have
\begin{eqna}
    T=\puoli m\int_0^\vec{v}\dd(\dvec{x}^2)=\puoli m \vec{v}^2.
\end{eqna}
This is called the kinetic energy of the particle. We can also write it as
\begin{eqna}
    T=\puoli mg_{\coa\cob}\vec{v}^\coa \vec{v}^\cob.
\end{eqna}
For the particle in the gravitational field this is just \(\puoli\dot{h}^2\) since we chose the unit of mass so that for that particle it is one.

\heb Obviously, if we have many particles the total kinetic energy is just the sum of individual kinetic energies, since we must accelerate each particle independently of the others. Labeling particles with \(i\), we have
\begin{eqna}
    T=\puoli\sum_i m_ig_{\coa\cob}\vec{v}_i^{\coa}\vec{v}_i^{\cob}
\end{eqna}
Now, we can take advantage of the tensor formalism we developed. The above sum is clearly linear in each of the velocities in it. If we have for example three particles, the sum is a linear function of six velocity vectors, that is, a tensor with six lower indices. We can therefore write
\begin{eqna}
    T=\puoli m_{\coa\cob\coc\cod\coe\cof}\vec{v}_1^{\coa}\vec{v}_1^{\cob}\vec{v}_2^{\coc}\vec{v}_2^{\cod}\vec{v}_3^{\coe}\vec{v}_3^{\cof}
\end{eqna}
where \(m\) is a tensor, mathematically the direct sum of all \(m_i g\)'s?, which takes care of calculating the kinetic energies of all the particles.

\heb And we can go even further! A you can see, we probably want to contract the velocities of all particles or none of them. Therefore denote three balls with just one and write
\begin{eqna}
    T=\puoli m_{\coa\cob}\dot{q}^\coa\dot{q}^\cob.
\end{eqna}
Here \(\dot{q}\) denotes collectively the velocities of all particles.

\hea Damn this formalism is beautiful.
\section{Configuration space}
\hea It looks like we have only one particle now. Funny.

\heb We can think that we really have only one particle, the `system particle', instead of the original \(N\) particles. The system particle moves in \(3N\)--dimensional so--called configuration space. Position in configuration space is often denoted by \(q\) and the velocity naturally by \(\dot{q}\).

\heb Configuration space is a manifold, but unlike the ordinary space, it does not have gyroscopes nor a natural metric. So we cannot parallel transport or change index positions, but that is not a great diffiiculty.

\hea Can we write Newton's second law for the system particle?

\heb Surely we can. For one particle it can be written as
\begin{eqna}
    \vec{F}=m\ddot{\vec{x}}\orr\vec{F}_\coa=mg_{\coa\cob}\ddot{\vec{x}}^\cob.
\end{eqna}
For the system particle we should obviously write
\begin{eqna}
    F_\coa=m_{\coa\cob}\ddot{q}^\cob,
\end{eqna}
where \(F\) contains all the forces of the individual particles.

We can also write
\begin{eqna}
    F=\dot{p}\wheree p_\coa=m_{\coa\cob}\dot{q}^\cob,
\end{eqna}
since the tensor \(m\) is constant, at least if the masses of the particles are constants.

\heb If we have only one particle, we have
\begin{eqna}
    \vec{p}_\coa=mg_{\coa\cob}\dvec{x}^\cob\orr\vec{p}=m\dvec{x}
\end{eqna}
and Newton's second law reads
\begin{eqna}
    \vec{F}=\dvec{p}.
\end{eqna}

\heb The quantity \(p\) is called momentum. It in some sense quantifies the amount of motion in the system: the greater the mass and the greater the velocity, the greater the momentum is. It is proportional to the velocity, and therefore is not an absolute vector like the acceleration is.% paranteluuu kaikenmoista

\heb In terms of momentum the second law can be read as `Force equals the rate of change of momentum,' that is, force changes the amount of motion.

\hea Sounds very reasonable.% liikkeess oleva voima voidaan ajatella paikallaan olevana voimana.?ehk?
\subsection{Potential energy}
\heb We are not the only ones who are able to produce force and do work. Mutual interactions of particles can often be thought of as arising from forces they excert on each other.

\heb Almost any large body can be thought of consisting of a large number of point particles excerting forces on each other. The forces keep the body in shape, but also may give rise for example for vibrations.

\hea So a complicated body can be represented as a point paricle in a large configuration space!

\heb Yes. The mutual forces of the particles appear as an external force acting on the system particle, like the gravitational force for the stone.

\heb The force in the configuration space depend on the state of motion of the system. I could for example depend on \(q\), \(\dot{q}\) and \(\ddot{q}\). However, here we have to restrict to a narrow class of forces, but don't worry, actually the forces we will restrict on are the most interesting forces.

\hea What forces we abandon?

\heb Right now let us abandon forces that depend on the changing of \(q\), that is, we consider forces that only depend on \(q\).

\heb For such a force \(F\) we can calculate how much work must be done to change the configuration from \(q=a\) to \(q=b\). The work \(W\) is
\begin{eqna}
    W(a\rightarrow b)=\int_{a}^{b}F\cdot\dd q.
\end{eqna}
This may in general depend on the path taken, but if it does not, we can form a function \(V\) of the configuration \(q\) by setting
\begin{eqna}
    V(q)=W(a\rightarrow q).
\end{eqna}
Now we have%pitaa tsekata merkkej
\begin{eqna}
    F\cdot\dd q=-\dd V=-\nabla V\cdot\dd q,
\end{eqna}
where \(\nabla V\) is the gradient of a \(V\), a vector that points points in the direction in which \(V\) grows most fast, its length of course depending on this rate.

\heb Newton's second law can therefore be written as
\begin{eqna}
    \dot{p}+\nabla V=0\orr m_{\coa\cob}\ddot{q}^\cob+\nabla_\coa V=0.
\end{eqna}
Looks simlpe, doesn't it!

\heb Now the kinetic energy can be written in a similar from.ldfcdlf
\begin{eqna}
    p_\coa=m_{\coa\cob}\dot{q}^\coa=\puoli\dot{\nabla}m_{\coa\cob}\dot{q}^\coa\dot{q}^\cob=\dot{\nabla}_\coa T.
\end{eqna}
Newton's second law becomes
\begin{eqna}
    \der{}{t}\dot{\nabla}T+\nabla V=0
\end{eqna}
or, after defining a new function \(L(q,\dot{q})=T-V\), the Lagranngian of the system,
\begin{eqna}
    \der{}{t}\dot{\nabla}L-\nabla L=0.
\end{eqna}
The equation of motion in this form is called the Lagrange's equation of motion.

\hea What can you say about the potential energy of our particle in a gravitational field?

\heb The force is constant, so it is almost trivial that it is the derivative of some constant \(\lambda\) times \(h\). The Lagrangian of the stone is therefore
\begin{eqna}
    L=\puoli\dot{h}^2-\lambda\,h.
\end{eqna}
\section{Action}
\heb The Lagrangian is actually very cool and useful function. Think about for example the vertically moving stone. If we neglect gravity, the Lagrangian is ridiculously simple, just
\begin{eqna}
    L=\puoli\dot{h}^2.
\end{eqna}
The equation of motion is very simple, isn't it?

\hea Well yes, there are no forces so Newton's second law holds and \(\ddot{h}=0\). The speed of the stone is constant.

\heb Along the motion the Lagrangian has some value at every moment, so we can condsider its time integral. It is called action of the motion and is denoted by \(S\). Clearly \(S\) the average value of the Lagrangian multiplied by the time of flight.

\heb Let's say we know something specific about the motion. We know the stone was at two particular moments \(t_a\) and \(t_b\) at at the heights \(h_a\) and \(h_b\).

\hea Ok. So from this information we immediately can calculate the velocity of the stone since we know the stone moves with a constant speed. And then we know also the average value of the Lagrangian.

\heb Now let us imagine that the stone did not move with a constant speed, but the motion still satisfies our boundary conditions. Such a motion is of course fictious, but we're equipped with an imagination. How does the action \(S\) change?

\hea Hmm. I don't know, it probably depends on our imagination on a very complicated way.

\heb But there is something that definitely happens. Think about the average of the speed.

\hea Aha! If the stone moves with a varying speed, it does some zig zag motion, and to get to the destination in time it must on average go faster than if it went uniformly. That means that the value of the action is bigger.

\heb So the realistic motion has the smallest. \(S\) compared to the fictious motions.

\heb if we now turn the gravity on, the Lagrangian gets a new term \(-\lambda h\). For simplicity now assume that \(h_a=h_b\). What can you say about the motion?

\hea It must first go up and then down, all in a smooth fashion. We solved this and the answer was a parabola.

\heb Now the avarage of the speed is not anymore the smallest of all imaginable motions. But the potential energy term is negative and compensates.

\hea Do you mean that by touring high in the potential the stone again manages to make the value of the action the lowest?

\heb At least that's what it looks like! If the stone visited at even higher altitudes that it actually does, it would probably make the average speed too high.

\hea I see, the parabola is probably the best compromise of having a fairly low average speed and a fairly high average potential at the same time.

\heb We can put this so--called principle off least action on a firm ground.
\subsection{Variation of the action}
\heb Let us consider any system with a Lagrangian \(L(q,\dot{q})\). The action of any motion \(q(t)\) is
\begin{eqna}
    S[q(t)]=\int L(q,\dot{q})\,\dd t.
\end{eqna}
The action can be thought of as a real--valued function in the space of all possible and impossible motions. We take a motion, divide it to small segments \(\dd t\), fsdf

\heb Now compare this action to the action of a slightly deformed path \(q(t)+\delta q(t)\). The action changes by
\begin{eqna}
    \delta S=\int_a^b\delta L\,\dd t.
\end{eqna}
The change in the Lagrangian of course depends on the moment of consideration \(t\) but I do not bother to write the time dependence always explicitly.

\heb Basic differential calculus tells us that
\begin{eqna}
    \delta L=\nabla L\cdot\delta q+\dot{\nabla}L\cdot\delta\dot{q}.
\end{eqna}
%The value of the Lagrangian at a specific moment of this path changes by
%\begin{eqna}
%    \delta L=\
All these are of course functions of time.

\hea What is \(\delta\dot{q}\)? I guess we can only specify \(\delta q\).

\heb It is of course determined by \(\delta q\). See
\begin{eqna}
    \delta\dot{q}=\der{(q+\delta q)}{t}-\der{q}{t}=\der{\delta q}{t}.
\end{eqna}

\hea Oh yes.

\heb It is exactly the same thing as the commuting of partial differentiation.

\heb We defined \(p=\dot{\nabla}L\), so now we have
\begin{eqnb}
    \delta L&=&\nabla L\cdot\delta q+p\cdot\der{}{t}\delta q\\
            &=&\nabla L\cdot\delta q+\der{}{t}(p\cdot\delta q)-\dot{p}\cdot\delta q\\
            &=&\der{}{t}(p\cdot\delta q)-(\dot{p}-\nabla L)\cdot\delta q.
\end{eqnb}
This holds for any motion \(q(t)\), were it realistic or not, and any small variation \(\delta q(t)\). But if \(q(t)\) is a real motion, it satisfies the equation of motion \(\dot{p}-\nabla L =0\) and therefore
\begin{eqna}
    \delta L=\der{}{t}(p\cdot\delta q).
\end{eqna}
This is a time derivative, so the change in \(S\) is just
\begin{eqna}
    \delta S=[p\cdot\delta q]_a^b.
\end{eqna}

\heb When we talked about the stone, we considered only motions with the same boundary conditions, right?

\hea Right.% Vuorovaikutustermit, ekvivalentit lagranget, using the action principle, 

\heb If we compare the realistic motion only to motions that begin and end at exactly same configuration, then \(\delta q(t_a)=\delta q(t_b)=0\) and therefore
\begin{eqna}
    \delta S=0.
\end{eqna}
So for a realistic motion that satisfies the equation of motion the action has a stationary value with respect to all small variations that vanish at the endpoints.

\hea Does this work also in the other direction?

\heb Yes it does, as can be easily seen. For any path and any small variation that vanishes at the endpoints we have
\begin{eqna}
    \delta =-\int_a^b(\dot{p}-\nabla L)\cdot\delta q\,\dd t.
\end{eqna}
If we know this vanishes, the stuff in the parenthesis must vanish at all times. If it did not do that, we could perform a variation which is nonzero only at for a short period at the right time and get \(\delta S\neq 0\). But the vanishing of the stuff in the parenthesis is just the equation of motion.

\hea So we have two equivalent characterizations of a realistic motion, that it satisfies the equation of motion and that it has a stationary value of the action with respect to small variations that vanish at the endpoints. Interesting.

\hea Why is this called the principle of \emph{least} action?

\heb Think of an ordinary real valued function which has a minimum point, for example \(y=x^2\). If we sit at \(x=0\) and perform a small variation \(\delta x\), the value of the function does not really change. In contrast, if we were sitting at \(x=1\), the value would have changed by \(\delta y=2\,\delta x\). The \(2\) is the derivative of \(x^2\) at \(x=1\).

\heb You can think of the action as a real--valued function in the space of all possible and impossible motions of the system. A real motion is a bottom of a kuoppa of \(S\) in the space of motions like \(x=0\) is the bottom of the \(x^2\) kuoppa on the real number line.

\hea Yes, I see, but if I had \(y=-x^2\), then I would also have \(\delta y=0\) even though there is no hole.

\heb That is an excellent notion. The stationary value can be a local minimum, a local maximum or a saddle point, in the fashion of \(y=x^3\). However, if you think of a realistic motion, you can always add a little zig zag to it, without really changing the potential energy but with increasing the speed. So a real motion cannot be a maximum at least if we have a Lagrangian of the form \(L=T-V\).

\heb So the the word `least' in the name of the principle is a small misnomer. A better name would be `the principle of stationary action'. But it really doesn't matter if we had a minimum or a saddle point; the essential thing is that the value is stationary.

\hea Okay. What if the Lagrangian depended also for example on \(\ddot{q}\)?

\heb If we had \(L=L(q,\dot{q},\ddot{q})\) and assumed that the equation of motion was the same Lagrange's equation, the action would not have had a stationary value. If we try to minimize such an action, we will find that the fourth--order differential equation
\begin{eqna}
    \nabla L-\der{}{t}\dot{\nabla}L+\der{^2}{t^2}\ddot{\nabla}L=0
\end{eqna}
must be satisfied by the motion, and even then \(\delta S=0\) would 

The equation reduces to Lagrange's equation if \(\ddot{\nabla}=0\).


in which the last term is uutta to the usual Lagrange's equation, were equivalent to the condition that \(\delta S=0\) with respect to all small variations with not only fixed endpoints but also fixed first time derivatives at the endpoints.

\heb This is because a fourth--order differential equation needs four boundary conditions, here the derivatives and configurations of the endpoints. The usual Lagrange's equation is of second order and there fore needs two boundary conditions, reflected to the action principle as the two fixed endpoints.
\subsection{Let us appreciate the action principle}
\heb The two equivalent views, the equation of motion and the action principle, are in some sense a local and a global perspectives to realistic motion of the system.

\hea Yeah, if I think of the equation of motion, I think that something happens at a specific instant of time. But the action principle evidently sees the motion in some sense as a whole.

\heb They are both good perspectives. If you need to actually find a concrete motion of a system, you better solve the equation of motion. But many theoretical aspects are more transparent when being analysed in terms of the action. The principle of least action gives some invaluable insights.

\heb It is good to pause and think of what we've done. You remember?

\hea In short we tried by common sense sort out some simple and reasonable equations that could describe motion oof particles. With a pretty good success we ended up with Newton's lawa, Lagrange's equation and the action principle.

\heb Essentially we noticed that motion of particles is smooth and therefore is describable by differential equations. Our exparience is that the future of a system depends on its initial velocity and configuration.

\hea So we ended up with a configuration space in which the motion is a function of time, a function which obeys a second order differential equation.

\heb Then we restricted the possible forces by assuming that the equation of motion can be written as
\begin{eqna}
    \der{}{t}\dot{\nabla}L-\nabla L=0.
\end{eqna}
That equation is equivalent with the action principle.

\heb The restriction forces us to abandon for example frictional forces, but I promise that very many interesting systems---I really mean that---can be described with an action.

\hea So if I'm trying to find a description for some physical system, you encourage me to look for a Lagrangian?

\heb Yes. If we want to describe a system, usually the best approach we can take is to look for a Lagrangian. In some situations it may be easier to find the action directly.

\heb Sometimes boundary conditions may be different and sometimes \(L\) may depend on \(\ddot{q}\). Or only on \(q\).

\hea Your instructions sound a bit muddy.

\heb There is no prescription. There is no road guaranteed to get you where you want. All I can say is that it is probably a good idea to look for some kind of action.
\subsection{Noether's theorem}
\heb If we can change the motion of a physical system in such a way that changed motion is also a realistic, the system is said to possess a symmetry. We can for example think of the motion of a thrown rock. We can move the throw to another city, or perhaps a hundred meters up, and get another possible motion of the rock.

\hea So you ar trying to say that the rock flies in the same way independently of the city and altitude where it has been thrown? I know that. You managed to choose pretty complicated words for expressing such a simple fact.

\heb Please excuse me. The flying rock is saide to possess a translation invariance. But the flight of the rock is an arc. If you turn it on its side, we get a flight which turns lef or right but not down---the rock clearly do not possess roational invariance around an axe parallel to ground.

\heb The translational invariance of the rock is a continuous symmetry, meaning that if we wish, we can do only a very small translation.

\heb A realistic motion, for example an arc of the rock, has two significant, though equivalent, properties: it satisfies the equation of motion and the value of the action is stationary for it. So, what quantifies our symmetry transformed motion?

\hea Obviously the same two properties.

\heb Exactly. Say we perform a small symmetry transformation \(\delta q\) to the motion. The value of the action remains stationary with respecto variations which vanish at the endpoints if the change in the action depends only on the endpoints, that is,
\begin{eqna}
    \delta S=f(A)-f(B)
\end{eqna}
for some \(f(q,\dot{q})\). This means that it must hold that \(\delta L=\dot{f}\). On the other hand, for any small change in a realistic motion we got
\begin{eqna}
    \delta L=\der{}{t}(p\cdot\delta q).
\end{eqna}
In totality we have
\begin{eqna}
    \dot{f}=\der{}{t}(p\cdot\delta q),
\end{eqna}
in other words the quantity
\begin{eqna}
    p\cdot\delta q-f
\end{eqna}
\subsection{Looking for an action}
\hea Well, if I'm after a Lagrangian, or an action, do you have any tips?

\heb First of all, if you take any action and multiply it by a number the stationary points obviously do not change. Also, if you take a function \(f=f(q)\) and add its time derivative to a Lagrangian, the stationary points remain unchanged.

\hea Why?

\hea The action changes by
\begin{eqna}
    S\rightarrow \int{}
\end{eqna}
% Numerolla  kertominen ja derivaatan lisminen ei muuta mitn

% Systeemien kytkeminen

% Symmetrioiden hydyntminen

% Hiriteoria

\heb If you have two independent systems, you can obviously 
\subsection{Constrained systems}
\hea I was thinking that if my system consist of for example two rocks connected by a rigid rod of length \(l\), then the rod constrains the motion. Is there a way to deal with the problem without giving any thoughts on the possibly complicated forces that keep the rod together?

\heb Yes there is. But lets first think about the actual forces that keep the rocks at a fixed distance at each other. The rod is probably perfectly rigid, but acts like a very, very stiff spring.

\hea Yeah, you're right. An infinitely stiff spring is a rigid rod.

\heb A spring has an equilibrium lengthm and if it is stretched or compressed, it generates a force pushing or pulling towards the equilibrium.

\hea Such a force can be obviously described by a potential which is zero at the equilibrium length and higher elsewhere.

\hea So let us subtract from the original Lagrangian of the two rocks a potential function which is zero when the distance of rocks is \(l\), and rises steeply when the distance gets longer or shorter. 

\heb I see. Now the Lagrangian implements the contstraining effect of the rock. If the potential has been chosen as rising very steeply, the stationary value of the action can only be achieved if the distance between the rocks is \(l\). Nice.

\hea Now, the action vanishes with respect to all variations of fixed enpoints, including the ones that does not change the distance \(l\) of the rocks.

\heb So we can forget the potential describing the forces of the rod and just find a stationary value of the original action not from all motions but from those that keep the distance between the stones \(l\). This is what I was after.

\hea But how can we achieve that?

\heb There is a trick for doing that. If we have a constraint like the rod in your example, it can be expressed as a function \(g(q)\) of the coordinates in such a way that \(g(q)=0\) exactly when the the constraint is satisfied. In the case of the roocks we could choose \(g\) to be the distance of the rocks minus \(l\).

\heb Now take the original Lagrangia and define a new one \(\primed{L}\) by \(\primed{L}=L+\lambda g(q)\) with \(\lambda\) an additional function of time. Now we perform the usual variation \(\delta q(t)\) \emph{and} a variation \(\delta \lambda(t)\). We get
\begin{eqna}
    \delta\primed{L}=\delta L+\lambda\partial g\delta q+g\delta\lambda.
\end{eqna}
delta s on jotain
If we vary only \(\lambda\), we get \(g=0\), which is just the constraint. If we vary \(q\) we get
\begin{eqna}
    \dot{p}-\partial L+\lambda \partial g.
\end{eqna}

%partial to nabla


% ALL symmetries of the flying stone


% Constrained motion / Lagrange's multipliers

% The range of applicability of the Lagrangian method


%In one dimension \(\partial_x f(x)\) is a dual vector, because \(\partial_x f\dd x=\dd f\) is a number.

\end{document}
\chapter{Coordinating things}
\section{Basis}
\section{Isometries}
\section{Groups}
\section{The Galilean group}
\section{A handful of particles}

\chapter{Vibrations}
\section{The simple armonic oscillator}
\section{Vibrations of several degrees of freedom}
\section{Fourier transformation}

\section{two particles}

\chapter{Ridiculously many particles}
\section{Phase space}
\section{ddf}


\newpage


\heb Let us consider some observers like you and me. Let's assume that the observers do not move and they all use same units of time and distance. How can these observers differ from each other?

\hea Well, I think everybody is different. Even if two persons were cloned from the same DNA, they have necessarily faced different things in life and become different kind of persons.

\heb Very clever. I wouldn't have known that! But you know that I meant something simpler.

\hea The observers are located at different places.

\heb And their noses may point in different directions. It is interesting to think of how the perspectives of these different observers are related to each other.

\heb We human beings have two eyes and therefore a stereo vision. We can locate things accurately in space by just watching. Say there is a couple of stones around and every observer see them. The question is: how are their points of view, I mean literally the image that gets projected on their verkkokalvo, relate to each other?

\hea Sounds like a complicated question.

\heb Well, the actual transformation of the view may be complicated, but we can say something about them. Let us denote the image on the verkkokalvos of some observer by \(\ket{\Psi}\). This notation is from quantum mechanics; I chose it because it looks so cool. Here it only means the view and nothing quantum mechanical.

\heb If we for example translate the viewer, that is change her location, the view changes. Let us denote the translation by \(\vec{T}\) and write
\begin{eqna}
    \ket{\Psi}\rightarrow\vec{T}\ket{\Psi}.
\end{eqna}

\hea Sounds trivial.

\heb I know. Now we can make another translation \(\primed{\vec{T}}\). In totality we have
\begin{eqna}
    \ket{\Psi}\rightarrow\vec{T}\ket{\Psi}\rightarrow\primed{\vec{T}}\vec{T}\ket{\Psi}.
\end{eqna}
Clearly both translations together form another translation which we may think of as a product \(\primed{\vec{T}}\vec{T}\). We can also think of doing nothing as a special kind of translation and denote it by \(1\).

\heb This is to say that mathematically the translations form a group. It is clear that the order of translations does not matter; we have \(\primed{\vec{T}}\vec{T}=\vec{T}\primed{\vec{T}}\) for any two translations. Translations are said to commute.

\hea I've heard of groups. If we translate by \(T\) we can always translate back, which I guess we could denote by \(T^{-1}\). It holds that \((\primed{T}T)^{-1}=T^{-1}\primed{T}^{-1}\) because
\begin{eqna}
    T^{-1}\primed{T}^{-1}\primed{T}T=T^{-1}1T=1.
\end{eqna}
This is clear because translating and then translating back is doing nothing. I feel stupid when I say that out loud.

\heb Good. Things get more interesting when we consider rotations. Rotations also clearly form a group which in three dimensions is known as \(SO(3)\).

\heb The group of rotations is more interesting because it is not commutative. Look: If I take this chair and do two \(\circc/4\) rotations around vertical and horizontal axes, I get different results depending on the order.

\hea I see, but what is \(\circc\)?

\heb It is the circle constant \(\circc\doteq2\pi\). \(\pi\) is used only for historical reasons; \(\circc\) is much more natural as it represens the whole circle. Check \url{tauday.com} for exhaustive discussion. If you don't like this, just go to the sourcecode of our lives and uncomment one line. The source can be found at \url{konstakurki.co/time}.

\hea Cool.






































\chapter{Curvature}
\heb If we take two parallel staight lines in space and extend them, do they remain parallel?

\hea Of course.

\heb How do you know that?

\hea Well, hmm, it seems obvious, doesn't it? I can't see how else could it be. Every straight lines that I've seen too be parallel at some point have always been parallel also everywhere else.

\heb OK, so you understand that it is an empirical observation. And then you of course also understand that it is not necessarily perfectly true, since we have really seen \emph{very} long straight parallel lines. Our picture of our space may be a bit too ideal

\hea Well, in principle yes, but this sounds like nitpicking to me. You said that usually the most simple guess is true, so I propose that parallel lines continue to be parallell since I think it is simpler than `parallel lines may stop being parallel'.

\heb That's surely what I did say. But what really is simple does not necessarily loook like that at first sight. This way or that, I really do have a point.

\heb Look at this karttapallo. Take two straight routes from the equator to the north pole. In the beginning they are parallel but they cross at the north pole and are definitely not parallel there.

\hea Ha ha. This really is stupid. The routes are of course really not straight; a straight route would go away from the surface of Earth to space. Earth is a ball in our three--dimensional space.

\heb Are you absolutely sure about that? Maybe we assume it is so because we have noticed that parallel lines do no continue as parallel and in reality our space is more peculiar that we assume.

\hea I think we can safely assume that light goes truly straight, and we know that a ship sailing away will eventually sink under the horizon. Earth really is a ball.

\heb Well, that is solid I must admit. Earth is a ball. But let me ask you another question. If we take an arrow, parallel transport it, that is transport it without changing the direction it points to, and parallel transport it back, does it end up pointing in the same direction to which it pointed in the beginning?

\hea I could of course ask how you know that the direction does not change, but I guess a gyroscope can be used to handle that issue. If we transport a gyroscope, hell yeah it will come back pointing at the original direction.

\heb Again, look at this model of Earth. Take an arrow and start from the equator, and go up to the north pole. Then go straigh right as long as the arrow is back on the equator. Then go back to the starting point along the equator. The arrow has turned one quarter of a full turn \(\circc\)!

\hea That's an interesting remark. But if you had really parallel transported the arrow, it would have pointed straight away from Earth at the north pole. You are again fooled by the fact that Earth is a ball.

\heb Yes I am. But now I tell you something that's gonna make you shit in your pants.

\heb Gyroscopes have actually been carried on satellites around Earth, and the results show that their directions \emph{do} change.

\hea What?

\heb Yes they do. If you don't believe, check the study of this and that. It can be found at arXiv under the number \arxivreference{1234.5678}.

\hea Don't worry, I believe you. But how is that possible?

\heb Well, that is something that will make us wonder a lot. But we may think that our space is actually intrinsically curved like the surface of Earth is.

\hea Is our space actually floating in some higher dimensional space?

\heb Maybe yes, maybe not. But it is actually not relevant for us, like the fact that Earth is embedded in a three--dimensional space is not relevant for a cartoghaper who just wants to draw useful two--dimensional pictures of land, seas and infrastructure built by human beings.

\heb We therefore are better to get used to curved spaces, or manifolds as they are often called in mathematics.






























\section{Riemannian manifolds}
\heb In mathematics a Riemannian manifold is a smooth space in which every curve has some length associated with it. Smooth means that the space is essentially flat if we look only small portions of it.

\heb For example the surface of Earth is a two--dimensional Riemannian manifold. It is smooth; if we take a small piece of this karttapallo it definitely is flat like a piece of paper. We can measure thee length of a curve for example by walking along it and counting the steps.

\heb But for example a cone is not smooth because the krki is sharp no matter how close we look and thus is is not a Riemannian mmanifold.

\hea Seems like an intuitive concept. I guess our space is a three--dimensional Riemannian manifold?

\heb The truth is actually even more facinating, but right now we regard it as such. In any case manifolds will be vital for us if we want to understand time.
\subsection{Parallel transport}
\subsection{Vectors}
\heb Think of two points on a manifold so close to each other that the manifold is essentially flat on the scale of their separation. We can draw an arrow from the first to second. Let me denote the arrow, which is naturally called the separation of the two points, by \(\dd x\).

\heb Now if there are other two points near the first ones, with a separation \(\dd y\), we can naturally add them by parallel transporting the other to start from the end of the other. We get another vector. Because the manifold is essentially flat in the region containing the vectors there is no ambiguity in carrying out the parallel transport.

\hea You're addoing vectors!

\heb Yes I am!

\hea We could also multiply \(\dd x\) by a number \(a\) producing a separation of two points along the same line as the originals but \(a\) times further apart of each other. And so on, yes, I know about vectors.

\heb The separation \(\dd x\) is naturally a vecor, but we can also have other vectors. For example if we have a particle moving in a manifold, we can consider its location at two different times separated by a small time interval \(\dd t\). If the particle does not move unreasonably fast, the locations at \(t\) and \(t+\dd t\) are close together. Dividing this separation \(\dd x\) by \(\dd t\) we get another arrow
\begin{eqna}
    \der{x}{t}\doteq u,
\end{eqna}
the velocity, which is also clearly a vector.

\hea But now \(\dd t\) is very short and so \(u\) might stretch so far that the space cannot anymore be considered as flat?

\heb Well, for general vectors we imagine the small essentially flat region to extend to infinity as perfectly flat. Imagine gluing a flat piece of paper on the karttapallo. In mathematics this flat piece is called a tangent space. If you are interested in mathematical details, consider for example the excellent book \emph{Geometry, Topology and Physics} by Mikio Nakahara.

\hea Ok, I'll check it. Now, if I'm correct, each point have its own tangent space. We must think of a vector to actually be located to somewhere in the manifold, and only vectors located at the same point, and belonging to the same tangent space, can be added, ultimately because it is not clear how to move vectors over long distances.

\heb Exactly, exactly. Now think of the inner product \(A\cdot B\) of two vectors \(A\) and \(B\).

\hea It measures how much \(A\) and \(B\) overlap. It is symmetric, \(A\cdot B=B\cdot A\), and linear in its both arguments. If \(A\) and \(B\) are perpendicular to each other, we have \(A\cdot B=0\).

\heb Yes. Now it is interesting to take the inner product of \(\dd x\) with itself. We can also do the same for \(a\dd x\), which is \(a\) times longer separation. We get
\begin{eqna}
    (a\dd x)\cdot(a\dd x)=a^2\,\dd x\cdot\dd x
\end{eqna}
because the inner product is linear in both of its arguments.

\hea I'm getting this! An inner product can be used to measure the distance of closely separated points.

\heb Yes. We can equip tangent spaces with any inner product we want. In Riemannian geometry the inner product is chosen so that
\begin{eqna}
    \dd s^2=\dd x\cdot\dd x
\end{eqna}
where \(\dd s\) is the distance of two points separated by \(\dd x\).

\heb The inner product is a bilinear map which takes two vectors and produces a number. It is useful to consider linear maps from any number of vectors to real number. Such objects are called tensors.

\heb The Riemannian inner product is usually denoted by \(g\). It is called the metric tensor, because it is used to measure distance. In mathematics literature one usually writes \(A\cdot B\doteq g(A,B)\), but this kind of notation gets complicated if we have tensors with more `inputs' for vectors. I like to write it as
\begin{eqna}
    A\cdot B\doteq g_{\coa\cob}A^\coa B^\cob.
\end{eqna}
Because the inner product is symmetric in its both arguments, we have \(g_{\coa\cob}=g_{\cob\coa}\).

\hea What are those colored balls?

\heb They are wireless connectors that connect the vectors to the inputs of the multilinear map \(g\). The position of the ball determines the input used and color determines to which it is connected.

\hea We can also think of a tensor \(T\) which take only one vector \(V\) and produces a number. That would be written as \(T_\coa V^\coa\). Tensors like \(T\) are called dual vectors.

\heb We can also think of the vector \(V\) as a linear map which takes \(T\) as its input. It really doesn't matter. That is why \(T\) is called a dual vector.

\heb We can also think of first taking an outer product \(T_\coa V^\cob\) and then connecting \(_\coa\) to \(^\cob\). It is easy to verify that the outer product is actually a bilinear map which which takes one vector and one dual vector as its inputs.

\hea I see. We can take tensors with any number of balls upstairs and downstairs, for example \(\tensay{F}{\coa}{\cob}\) and \(\tensya{K}{\coc}{\cod}\). Then we can form the outer product \(\tensay{F}{\coa}{\cob}\tensya{K}{\coc}{\cod}\), which takes inputs exactly as you would expect from the balls.

\heb Yes.

\hea Then we can connect balls, for example \(\tensay{F}{\coa}{\cob}\tensya{K}{\coc}{\coa}\).

\heb Connecting the balls is often called contracting. Mathematically it means taking a trace. An upper ball must always be contracted with a lower one and vice versa.

\hea And the contracted indices are not anymore inputs! A contraction takes one input and one output off a tensor.

\heb finally, when there are no iputs and out puts left, we have just a number. How simple is that?

\hea Very simple and elegant.

\heb In this notation we have
\begin{eqna}
    \dd s^2=g_{\coa\cob}\dd x^\coa\dd x^\cob.
\end{eqna}
\subsection{Length of a curve}
Now, think of a curve \(x(t)\) on a manifold. We can calculate its length \(s\) by thinking it consisting of very many small segments. We have
\begin{eqna}
    s=\int\dd s=\int\sqrt{g_{\coa\cob}\dd x^\coa\dd x^\cob}.
\end{eqna}
\section{Differentiation}
\hea Could we differentiate things?

\heb If we have a scalar function \(\phi(x)\)on a manifold, that is, we attach a number to each point in the manifold, it is easy to compare neighboring values: just take the difference \(\dd\phi\). If we have a curve \(x(t)\), we can easily calculate the derivative of \(\phi\) with respect to \(t\); it is just \(\dd\phi/\dd t\).

\heb But differentiating a vector field is more difficult because we have to compare vectors belonging to different tangent spaces. How could we do it?

\hea Well, the only natural way of doing it that comes to my mind is of parallel transporting one of the vectors near the other and then doing the comparison.
\section{Differentiation}
\heb If we have tensors defined on a curve or perhaps everywhere on the manifold, it would be nice to be able to differentiate them. But there is a problem: we have to compare neighboring vectors that belong to different tangent spaces. How could we do it?

\hea Well, if we compare vectors the only thing that comes to my mind is of parallel transporting one of the vectors near the other and then doing the comparison. But I don't know what to do to other kinds of tensors.

\heb Damn you're clever. The difference of a vector field \(A(x)\) gotten that way is usually denoted by \(DA\). The capital d is used to emphasize that it is not a simple subtraction of two numbers. Now, if we have a curve \(x(t)\), we can defferentiate \(A\) along it by forming the simple fraction
\begin{eqna}
    \frac{DA}{\dd t}\doteq\cder{A}{t}.
\end{eqna}
It is also a vector. We usually write \(D\) instead of \(\dd\) also before \(t\) because mixing two different d's looks stupid.

\hea But what about other tensors?

\heb Oh yes. Actually they can also be naturally parallel transported, meaning that if the original tensor gives some number out of some arbitrary vectors, the parallel transported tensor gives the same number out of the same vectors also parallel transported to the same location.

\hea Like always, the covariant derivative obeys the Leibniz rule.







\end{document}
